---
title: "Final project presentation 3"
author: "sadia kana"
date: "2023-08-14"
output: pdf_document
---

## Final project presentation 3

```{r}
# Load the necessary libraries
library(tidyverse)
library(reshape2)
library(wesanderson)
library(harrypotter)
library(ggpubr)
library(tagger)

library(tidyr)
library(dplyr)
library(pheatmap)
library(knitr)
library(kableExtra)
library(gganimate)
library(gapminder)
library(gifski)

```


```{r}
# Load the gene expression data and metadata
GeneExpression_data<- read.csv(file = "/Users/Student1/Downloads/QBS103_finalProject_geneExpression.csv")
Metadata<- read.csv(file =  "/Users/Student1/Downloads/QBS103_finalProject_metadata.csv", stringsAsFactors = F, na.strings = ("unknown"))

Metadata<- droplevels(na.omit(Metadata))

# Loak at both the csv files using head function

  head(GeneExpression_data)
  head(Metadata)

```


```{r}
# Rename the first column of GeneExpression_data
colnames(GeneExpression_data) [1]<- "Genes"

# Check the renaming

  head(GeneExpression_data)
```


```{r}
# Convert the gene expression data into long format
longData<- reshape2::melt(GeneExpression_data, id.vars = ("Genes"),
                          value.name = "gene_expression", variable.name = "participant_id")

# Check the long format conversion

  head(longData)

```


```{r}
# Link longData with Metadata

LinkedData<- merge(longData, Metadata, by = "participant_id")

# View the linked data

  head(LinkedData)
  tail(LinkedData)
  
```

# Generate a table of summary statistics for all the covariates you looked at, stratifying by one of your categorical variables. Tables should report n (%) for categorical variables and mean (sd) or median [IQR] for continuous variables. (10 pts).

```{r}


```

# Generate a heatmap of at least 10 genes with tracking bars for your two selected categorical covariates. Heatmaps should include clustered rows and columns. (10 pts)

```{r}
genelist <- LinkedData$Genes[1:15]

variance<- apply(LinkedData, MARGIN = 1, FUN = var)

# Order rows of miRNA so that highest variance in expression is on top
Data <- LinkedData[order(variance,decreasing = T),]

# Log2-normalize data for plotting
log2.data <- log2(Data)

pheatmap(log2.data[1:20,],
         cluster_rows = F,
         cluster_cols = F)

```

#Going through the documentation for ggplot2, generate a plot type that we did not previously discuss in class that describes your data in a new and unique way (5 pts)

```{r}
Gene_data<- LinkedData[LinkedData$Genes == "CD24", ]
Gene_data$ferritin.ng.ml.<- na.omit(as.numeric(Gene_data$ferritin.ng.ml.))
table(Gene_data$ferritin)

g<- ggplot(Gene_data, aes( x = ferritin.ng.ml., y =gene_expression))+
  geom_hex(bins = 30)+
  theme_classic()+
  labs(title =substitute(paste( "The association of ", italic("BPI")," ", "expression with ferritin")), x = "Ferritin", y = "CD24 expression")+
  theme(text = element_text(size = 11)) + 
  theme(plot.title = element_text(hjust = .35))
plot(g)
```

  
  
  